Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-12-11 paper.tex
No State monad
Reviewer A:
  p2 l54: “cryptographers have been programming in the State monad” je
  trouve cette affirmation un brin grandilocante quand-même. D'abord,
  il n'y a plus de monade dans le reste de l'article. En plus, si on
  peut effectivement décrire la composition de fonction $A\rightarrow
  A$ comme “programmer dans la monade d'état”, il faut tout de même un
  petit peu de mauvaise fois. En plus, c'est contreproductif pour
  l'argument que la crypto c'est de la programmation fonctionnelle,
  parce qu'il me semble que la composition de fonction, c'est plus ce
  à quoi on pense

Looking at the process of turning the plaintext into a ciphertext, one
realizes that cryptographers have been programming in the State monad
all along: ciphers are described as operating on a ``state'',
initialized to be the plaintext and read off at the end as the
ciphertext. This state is updated by a ``round'' function that is
repeated a given number of times (of the order of ten iterations). The
exact number of iterations is a trade-off between latency (less
iterations means faster computation) and security (more iterations
makes for more costly cryptanalysis effort). The round function itself
is built compositionally from 3 components: a key mixing layer (making
the encryption process reversible), a confusion layer (smoothing local
correlations among the inputs) and a diffusion layer (spreading
correlations across the whole output block).

>>>>

Ciphers are described as operating on a ``state'', initialized to be
the plaintext and read off at the end as the ciphertext. This state is
updated by a ``round'' function that is repeated a given number of
times (of the order of ten iterations). The exact number of iterations
is a trade-off between latency (less iterations means faster
computation) and security (more iterations makes for more costly
cryptanalysis effort). The round function itself is built
compositionally from 3 components: a key mixing layer (making the
encryption process reversible), a confusion layer (smoothing local
correlations among the inputs) and a diffusion layer (spreading
correlations across the whole output block).
